Value-Suppressing Uncertainty Maps

### 2017-08-22 paper/main.tex
Moved paper to CHI format

% ---------------------------------------------------------------------
% EG author guidelines plus sample file for EG publication using LaTeX2e input
% D.Fellner, v1.17, Sep 23, 2010

\title[EG \LaTeX\ Author Guidelines]%
      {\LaTeX\ Author Guidelines for EUROGRAPHICS Proceedings Manuscripts}

% for anonymous conference submission please enter your SUBMISSION ID
% instead of the author's name (and leave the affiliation blank) !!
\author[D. Fellner \& S. Behnke]
       {D.\,W. Fellner\thanks{Chairman Eurographics Publications Board}$^{1,2}$
        and S. Behnke$^{2}$
%        S. Spencer$^2$\thanks{Chairman Siggraph Publications Board}
        \\
% For Computer Graphics Forum: Please use the abbreviation of your first name.
         $^1$TU Darmstadt \& Fraunhofer IGD, Germany\\
         $^2$Institut f{\"u}r ComputerGraphik \& Wissensvisualisierung, TU Graz, Austria
%        $^2$ Another Department to illustrate the use in papers from authors
%             with different affiliations
       }

% ------------------------------------------------------------------------

% if the Editors-in-Chief have given you the data, you may uncomment
% the following five lines and insert it here
%
% \volume{27}   % the volume in which the issue will be published;
% \issue{1}     % the issue number of the publication
% \pStartPage{1}      % set starting page

%-------------------------------------------------------------------------
\begin{document}

% \teaser{
%  \includegraphics[width=\linewidth]{eg_new}
%  \centering
%   \caption{New EG Logo}
% \label{fig:teaser}
% }

\maketitle

\begin{abstract}
   The ABSTRACT is to be in fully-justified italicized text, 
   between two horizontal lines,
   in one-column format, 
   below the author and affiliation information. 
   Use the word ``Abstract'' as the title, in 9-point Times, boldface type, 
   left-aligned to the text, initially capitalized. 
   The abstract is to be in 9-point, single-spaced type.
   The abstract may be up to 3 inches (7.62 cm) long. \\
   Leave one blank line after the abstract, 
   then add the subject categories according to the ACM Classification Index 
   (see http://www.acm.org/class/1998/).

\begin{classification} % according to http://www.acm.org/class/1998/
\CCScat{Computer Graphics}{I.3.3}{Picture/Image Generation}{Line and curve generation}
\end{classification}

\end{abstract}

%-------------------------------------------------------------------------
\section{Introduction}

Please follow the steps outlined in this document very carefully when
submitting your manuscript to Eurographics.

You may as well use the \LaTeX\ source as a template to typeset your own
paper. In this case we encourage you to also read the \LaTeX\ comments
embedded in the document.

%-------------------------------------------------------------------------
\section{Instructions}

Please read the following carefully.

%-------------------------------------------------------------------------
\subsection{Language}

All manuscripts must be in English.

%-------------------------------------------------------------------------
\subsection{Margins and page numbering}

All printed material, including text, illustrations, and charts,
must be kept within a print area %6.31 inches (16.03 cm) 
7 inches (17.7 cm) wide by
%9.10 inches (23.13 cm) 
9.44 inches (24 cm) high. Do not write or print anything
outside the print area. Number your pages on odd sites right
above, on even sites left above, no page number on the first site.
Do not use page numbering within the final version of your paper.

%------------------------------------------------------------------------
\subsection{Formatting your paper}

% All text with the exception of the abstract must be in a two-column format.
% The total allowable width of the text area -- including header and footer
% lines -- is 161\,mm (6.34 inch) wide by 231\,mm (9.10 inch) high.
% 
% Columns are to be 76\,mm (3.0 inch) wide, with a 8\,mm (0.315 inch) space 
% between them.

All text with the exception of the abstract must be in a two-column format.
The total allowable width of the text area -- including header and footer
lines -- is 177\,mm (7 inch) wide by 245\,mm (9.64 inch) high.

Columns are to be 84\,mm (3.3 inch) wide, with a 8\,mm (0.315 inch) space 
between them.

The space between the header line and the first line of the text body and
between the last line of the text body and the footer line is 5\,mm
(0.196 inch) each.

%-------------------------------------------------------------------------
\subsection{Type-style and fonts}

Wherever Times is specified, Times Roman may also be used. If
neither is available on your word processor, please use the font
closest in appearance to Times that you have access to. Only
Type-1 fonts will be accepted.

MAIN TITLE. The title should be in Times 17-point, boldface type and
centered. Capitalize the first letter of nouns, pronouns, verbs, adjectives,
and adverbs; do not capitalize articles, coordinate conjunctions, or
prepositions (unless the title begins with such a word). Leave two blank
lines after the title.

AUTHOR NAME(s) and AFFILIATION(s) are to be centered beneath the title and
printed in Times 9-point, non-boldface type. This information is to be
followed by two blank lines.

The ABSTRACT is to be in a one-column format. The MAIN TEXT is to be in a
two-column format.

MAIN TEXT. Type main text in 9-point Times, single-spaced. Do \emph{not} use
double-spacing. All paragraphs should be indented 1 em (the length of the
dash in the actual font). Make sure your text is fully justified -- that is,
flush left and flush right. Please do not place any additional blank lines
between paragraphs. Figure and table captions should be 9-point Times
boldface type as in Figure~\ref{fig:firstExample}.

\noindent Long captions should be set as in Figure~\ref{fig:ex1} or
Figure~\ref{fig:ex3}.

\begin{figure}[htb]
   % an empty figure just consisting of the caption lines
   \caption{\label{fig:ex1}
     'Empty' figure only to serve as an example of long caption requiring 
     more than one line. It is not typed centered but aligned on both sides.}
\end{figure}

\noindent
Figures which need the full textwidth can be typeset as Figure~\ref{fig:ex3}.

\noindent Callouts should be 9-point Times, non-boldface type. Initially
capitalize only the first word of section titles and first-, second-, and
third-order headings.

FIRST-ORDER HEADINGS. (For example, \textbf{1. Introduction}) should be Times
9-point boldface, initially capitalized, flush left, with one blank line
before, and one blank line after.

SECOND-ORDER HEADINGS. (For example, \textbf{2.1. Language}) should be Times
9-point boldface, initially capitalized, flush left, with one blank line
before, and one after. If you require a third-order heading (we discourage
it), use 9-point Times, boldface, initially capitalized, flush left, preceded
by one blank line, followed by a period and your text on the same line.

The headline \emph{(authors / title)} must be shortened if it uses the full 
two column width of the main text.
There must be enough space for the page numbers. Please use ``et al.'' if 
there are more than three authors and specify a shortened version for your title.
%-------------------------------------------------------------------------
\subsection{Footnotes}

Please do \emph{not} use footnotes at all!

%-------------------------------------------------------------------------
\subsection{References}

List all bibliographical references in 9-point Times, single-spaced, at the
end of your paper in alphabetical order. When referenced in the text, enclose
the citation index in square brackets, for example~\cite{Lous90}. Where
appropriate, include the name(s) of editors of referenced books.

For your references please use the following algorithm:
\begin{itemize} 
\item \textbf{one} author: first 3 chars plus year -- 
      e.g.\ \cite{Lous90}
\item \textbf{two}, \textbf{three} or \textbf{four} authors: first char
      of each family name plus year --  e.g.\ \cite{Fellner-Helmberg93} 
      or \cite{Kobbelt97-USHDR} or \cite{Lafortune97-NARF}
\item \textbf{more than 4} authors: first char of family name from 
      first 3 authors followed by a '*' followed by the year -- 
      e.g.\ \cite{Buhmann:1998:DCQ} or \cite{FolDamFeiHug.etal93} 
\end{itemize}

For BibTeX users a style file \ \texttt{eg-alpha.bst} \ is available which
uses the above algorithm.

%-------------------------------------------------------------------------
\subsection{Illustrations, graphs, and photographs}

All graphics should be centered.

%%%
%%% Figure 1
%%%
\begin{figure}[htb]
  \centering
  % the following command controls the width of the embedded PS file
  % (relative to the width of the current column)
  \includegraphics[width=.8\linewidth]{sampleFig}
  % replacing the above command with the one below will explicitly set
  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
  % It will also prevent LaTeX from reading the PS file to determine
  % the bounding box (i.e., it will speed up the compilation process)
  % \includegraphics[width=.95\linewidth, bb=39 696 126 756]{sampleFig}
  %
  \parbox[t]{.9\columnwidth}{\relax
           For all figures please keep in mind that you \textbf{must not}
           use images with transparent background! 
           }
  %
  \caption{\label{fig:firstExample}
           Here is a sample figure.}
\end{figure}

If your paper includes images, it is very important that they are of
sufficient resolution to be faithfully reproduced.

To determine the optimum size (width and height) of an image, measure
the image's size as it appears in your document (in millimeters), and
then multiply those two values by 12. The resulting values are the
optimum $x$ and $y$ resolution, in pixels, of the image. Image quality
will suffer if these guidelines are not followed.

Example 1: 
%
An image measures 50\,mm by 75\,mm when placed in a document. This
image should have a resolution of no less than 600 pixels by 900
pixels in order to be reproduced faithfully.

Example 2: 
%
Capturing a screenshot of your entire $1024 \times 768$ pixel display
monitor may be useful in illustrating a concept from your research. In
order to be reproduced faithfully, that $1024 \times 768$ image should
be no larger than 85 mm by 64 mm (approximately) when placed in your
document.

%-------------------------------------------------------------------------
\subsection{Color}

\textbf{Please observe:} as of 2003 publications in the proceedings of the
Eurographics Conference can use color images throughout the paper. No
separate color tables are necessary.

However, workshop proceedings might have different agreements! 
Figure~\ref{fig:ex3} is an example for creating color plates.

%------------------------------------------------------------------------
\subsection{Embedding of Hyperlinks / Typesetting of URLs}

Due to the use of the package \texttt{hyperref} the original behavior
of the command $\backslash$\texttt{url} from the package \texttt{url}
is not available. To circumvent this problem we either recommend to
use the command $\backslash$\texttt{httpAddr} from the 
included package \texttt{egweblnk} (see below) or to replace the
command $\backslash$\texttt{url} by the command $\backslash$\texttt{webLink} 
-- e.g. in cases where $\backslash$\texttt{url} has been used
widely in BibTeX-References. In the latter case we suggest to run
BibTeX as usual and then replace all occurences of $\backslash$\texttt{url}  by
$\backslash$\texttt{webLink}

\noindent
The provided commands for hyperlinks, in a nutshell, are:

\begin{description} \itemsep 1ex
\item [\webLinkFont $\backslash$httpAddr \{URL without leading 'http:'\}]
      \mbox{}\\
      e.g. \  \httpAddr{//diglib.eg.org/EG/DL/WS}

\item[\webLinkFont $\backslash$ftpAddr \{URL without leading 'ftp:'\}]
      \mbox{}\\
      e.g. \  \ftpAddr{//www.eg.org/EG/DL/ftpupload}

\item[\webLinkFont $\backslash$URL \{url\}]
      \mbox{}\\
      e.g. \  \URL{http://www.eg.org/EG/DL/WS}

\item[\webLinkFont $\backslash$MailTo \{Email addr\}]
      \mbox{}\\
      e.g. \  \MailTo{publishing@eg.org}

\item[\webLinkFont $\backslash$MailToNA \{emailName\}\{@emailSiteAddress\}]
      \mbox{}\\
      e.g. \  \MailToNA{publishing}{@eg.org}

\item[\webLinkFont $\backslash$webLink\{URL without hyperlink creation\}]
      \mbox{}\\
      e.g. \  \webLink{http://www.eg.org/some_arbitrary_long/but_useless/URL}

\end{description}

%------------------------------------------------------------------------
\subsection{PDF Generation}

Your final paper should be delivered as a PDF document with all typefaces
embedded. \LaTeX{} users should use \texttt{dvips} and \texttt{ps2pdf} to
create this PDF document. Adobe Acrobat Distiller may be used in place of
\texttt{ps2pdf}.

Adobe PDFWriter is \emph{not} acceptable for use. Documents created with
PDFWriter will be returned to the author for revision. \texttt{pdftex} and
\texttt{pdflatex} (and its variants) can be used only if the author can
make certain that all typefaces are embedded and images are not downsampled
or subsampled during the PDF creation process.

Users with no access to these PDF creation tools should make available a
PostScript file and we will make a PDF document from it.

The PDF file \emph{must not} be change protected.

%------------------------------------------------------------------------
\subsubsection*{Configuration Notes: dvips / ps2pdf / etc.}

\noindent
\texttt{dvips} should be invoked with the \texttt{-Ppdf} and \texttt{-G0}
flags in order to use Type 1 PostScript typefaces:

\begin{verbatim}
    dvips -t a4 -Ppdf -G0 -o my.ps my.dvi
\end{verbatim}

\noindent
If you are using version 7.x of GhostScript, please use the following method of invoking \texttt{ps2pdf}, in
order to embed all typefaces and ensure that images are not downsampled or subsampled in the PDF
creation process:

\begin{verbatim}
  ps2pdf -dMaxSubsetPct=100 \
         -dCompatibilityLevel=1.3 \
         -dSubsetFonts=true \
         -dEmbedAllFonts=true \
         -dAutoFilterColorImages=false \
         -dAutoFilterGrayImages=false \
         -dColorImageFilter=/FlateEncode \
         -dGrayImageFilter=/FlateEncode \
         -dMonoImageFilter=/FlateEncode \
         mypaper.ps mypaper.pdf
\end{verbatim}

If you are using version 8.x of GhostScript, please use this method in place of the example above:
\begin{verbatim}
  ps2pdf -dPDFSETTINGS=/prepress \
         -dCompatibilityLevel=1.3 \
         -dAutoFilterColorImages=false \
         -dAutoFilterGrayImages=false \
         -dColorImageFilter=/FlateEncode \
         -dGrayImageFilter=/FlateEncode \
         -dMonoImageFilter=/FlateEncode \
         -dDownsampleColorImages=false \
         -dDownsampleGrayImages=false \
         mypaper.ps mypaper.pdf
\end{verbatim}

%------------------------------------------------------------------------
\subsubsection*{Configuration Notes: pdftex / pdflatex / etc.}

\noindent
Configuration of these tools to embed all typefaces can be accomplished by editing the \texttt{updmap.cfg} file
to enable inclusion of the standard (or base) 14 typefaces.

Linux users can run the \texttt{updmap} script to do this:
\begin{verbatim}
updmap --setoption pdftexDownloadBase14 true
\end{verbatim}

Windows users should edit the \texttt{updmap.cfg} files found in their TeX installation directories (one or both
of the following may be present):
\begin{verbatim}
  INSTALLDIR\texmf\web2c\updmap.cfg
  INSTALLDIR\localtexmf\miktex\config\updmap.cfg
\end{verbatim}

Ensure the value for \texttt{pdftexDownloadBase14} is "true," and then follow the instructions found here:
\httpAddr{//docs.miktex.org/manual/} to update your MikTeX installation.

%------------------------------------------------------------------------
\subsubsection*{Configuration Notes: Acrobat Distiller}

We recommend to download and install the version of the ``CMW'' Adobe Acrobat Distiller job options file
appropriate for your operating system and version of Acrobat from the following URL:

\httpAddr{//www.cadmusmediaworks.com/index2.html}\\
in the ``(Operating System)/Applications/Distiller Settings'' folder. The ``CMW'' job options file embeds
all typefaces and does not downsample or subsample images when creating the PDF document.
%------------------------------------------------------------------------
\subsection{Copyright forms}

You must include your signed Eurographics copyright release form
when you submit your finished paper. We MUST have this form before
your paper can be published in the proceedings.

%-------------------------------------------------------------------------
\subsection{Conclusions}

Please direct any questions to the production editor in charge of
these proceedings.

%-------------------------------------------------------------------------

%\bibliographystyle{eg-alpha}
\bibliographystyle{eg-alpha-doi}

\bibliography{egbibsample}

%-------------------------------------------------------------------------
\newpage

\begin{figure*}[tbp]
  \centering
  \mbox{} \hfill
  % the following command controls the width of the embedded PS file
  % (relative to the width of the current column)
  \includegraphics[width=.3\linewidth]{sampleFig}
  % replacing the above command with the one below will explicitly set
  % the bounding box of the PS figure to the rectangle (xl,yl),(xh,yh).
  % It will also prevent LaTeX from reading the PS file to determine
  % the bounding box (i.e., it will speed up the compilation process)
  % \includegraphics[width=.3\linewidth, bb=39 696 126 756]{sampleFig}
  \hfill
  \includegraphics[width=.3\linewidth]{sampleFig}
  \hfill \mbox{}
  \caption{\label{fig:ex3}%
           For publications with color tables (i.e., publications not offering
           color throughout the paper) please \textbf{observe}: 
           for the printed version -- and ONLY for the printed
           version -- color figures have to be placed in the last page.
           \newline
           For the electronic version, which will be converted to PDF before
           making it available electronically, the color images should be
           embedded within the document. Optionally, other multimedia
           material may be attached to the electronic version. }
\end{figure*}

\end{document}

>>>>

\section{Introduction}
certainty is an inescapable component of collecting, presenting, and using data. A common goal in the communication of uncertainty is promoting \emph{uncertainty-aware decisions}: the audience should be aware of the risks and rewards of certain decisions, modulate their confidence in their conclusions, and perhaps refrain from making a decision at all if there is too much uncertainty.  A way that designers can contribute to this goal is by ensuring that uncertainty information is \emph{well-integrated} with the rest of the data. That is, it should be difficult to discount or ignore the uncertainty in a dataset.

Simultaneous presentation of uncertainty and value necessitates the construction of a bivariate map\,---\,a relation, in terms of visual variables, between 2-tuples $(\text{value}, \text{uncertainty})$ and mark properties. The design of bivariate maps is difficult and, due to the interference and interplay between different visual variables, often leads to maps with few discriminable categories.

In this paper, we contribute \textbf{Value-Suppressing Uncertainty Maps} (VSUMs) for integrating data and uncertainty information in visualizations. VSUMs intentionally alias together data values with high uncertainty, affording greater discriminability as uncertainty decreases. Compared to a traditional two-dimensional bivariate map, with differing outputs for each combination of value and uncertainty, VSUMs can be thought of as arcs: as uncertainty increases, values are mapped to smaller and smaller sets of outputs, culminating in a singularity where all inputs are mapped to an identical, highly uncertain mark regardless of their data value. \figref{fig:example} shows examples of both a more traditional bivariate map and a VSUM.

We describe the motivations behind the use of VSUMs, examples of their utility for decision-making under uncertainty, and assess VSUMs in a crowdsourced experiment. Our experimental results indicate that VSUMs create close integration between uncertainty and data, promoting rapid but cautious decision-making.

\exampleFig
\section{Related Work}

% What this section needs to do:
% State of the art in uncertainty vis, with heavy emphasis on MacEachren & co.
% Bivariate maps are hard!
% Binning colormaps is de rigueur
% The variables we'd want to use for uncertainty (like value or alpha or size) mess with color discriminability

Despite the acknowledged importance of uncertainty in understanding data, explicit representations of uncertainty are often missing from visualizations~\cite{boukhelifa2009uncertainty}. This is partially due to the complexity of uncertainty as a concept. In typologies from Thomson \ea~\cite{thomson2005typology} and Buttenfield \& Beard~\cite{buttenfield1994graphical}, the authors note that many, occasionally contradictory, concepts can fall under the category of ``uncertainty,'' including data quality, sampling error, credibility, and provenance.

Reviewing the state of the art in uncertainty visualization in the field, Greithe \ea~\cite{griethe2006visualization} and Brodlie \ea~\cite{brodlie2012review} present lists of potential techniques for conveying uncertainty in concert with data. Some of these options (interaction, animation, and sonification) are not applicable for static charts; even for dynamic charts, many users do not interact with charts in sufficient detail to recover uncertainty information~\cite{nyt2016}. While we acknowledge the potential utility of these techniques in uncertainty visualization (for instance, the animation used to convey sampling error in Hullman \ea's Hypothetical Outcome Plots~\cite{hullman2015hypothetical}), we limit the scope of our discussion to techniques which center on static charts.

There are a number of techniques for communicating values and uncertainty in one-dimensional statistical graphics~\cite{olston2002visualizing, wickham201140, neyman1937outline}. Techniques such as error bars, confidence bands, and violin plots~\cite{hintze1998violin} use a separate position channel to encode uncertainty. The techniques in this paper also apply to 2D visualizations such as heatmaps. To our knowledge, there are no established standards of visualizing uncertainty in heatmaps.

\strategyFig

To support our goal of making uncertainty information well-integrated with data, we focus on the techniques of \emph{juxtaposition} (where uncertainty is presented alongside the data), \emph{superposition} (where additional glyphs are overlaid on top of the data), and \emph{explicit encoding} (where an additional visual channel is used to encode the uncertainty). Gleicher \ea~\cite{gleicher2011visual} explore these three strategies from the perspective of affording \emph{comparisons}; here, we examine them from the perspective of incorporating measures of uncertainty. \figref{fig:strategies} shows simple examples of each of these three techniques for presenting both uncertainty and value.

\subsection{Strategies for Encoding Uncertainty}

\emph{Juxtaposition} creates two explicit visualizations: an ``uncertainty map'' and a ``data map.'' MacEachren~\cite{maceachren1992visualizing} portrays uncertainty visualization as the design problem of how to \emph{unify} these two maps, and so this explicit separation can be counterproductive to the goal of information fusion. Juxtaposition also creates an additional search task for viewers, who must register hotspots in one map with corresponding features of the other map. Rather than complete this task, viewers may choose to ignore the uncertainty information altogether. This ignoreability is a key deficiency for information fusion tasks. Nevertheless, especially in cases where the quantification of uncertainty is a separate analysis process from the measuring of data values, juxtaposition is a common design tactic for conveying uncertainty~\cite{moritz2017trust}.

\emph{Superposition}, or the overlay of glyphs representing uncertainty on top of a data map, is another common strategy. Error bars are a common example of this strategy: a mark indicating measurement error is superimposed atop a mark representing a point value. Ware's~\cite{ware2009quantitative} \emph{textons} are an example of this strategy applied to thematic maps: monochrome glyphs that can encode an ordinal variable, or a binned quantitative variable. Greithe \ea~\cite{griethe2006visualization} and MacEachren~\cite{maceachren1992visualizing,maceachren1998visualizing} present other examples of superimposed uncertainty in thematic maps, including overlaid grids with ``fuzzy'' lines to indicate uncertainty, or overlaid treemaps with coarser and coarser resolution in uncertain regions. Costs to this approach include obscuring the underlying data map (which can cause uncertainty information to dominate the display~\cite{brodlie2012review}), potential visual interactions that can impair the legibility of glyphs (for instance, simultaneous contrast effects), and the necessity of performing spatial binning to ensure that there are a discrete number of uncertainty glyphs. How this binning is performed can introduce undesirable artifacts on the resulting visualization, suppressing important signals or highlighting spurious ones~\cite{battersby2016shapes}.

\emph{Explicit encoding} directly represents uncertainty with a visual channel, simultaneously with data. VSUMs employ this strategy. In addition to overcoming the drawbacks of the other techniques mentioned in this section, we believe that explicit encoding represents the most general form of the solution of how to fuse uncertainty information with data. The most pressing design decisions for explicit encoding are 1) which visual variables to use to encode uncertainty and 2) how to design bivariate maps which afford the encoding of both value and uncertainty. We deal with these two questions in greater detail below.

\subsection{Visual Variables for Representing Uncertainty}

The decision to explicitly encode uncertainty increases the dimensionality of the data, and so requires the use of (at least one) additional visual channel ~\cite{brodlie2012review}. Uncertainty therefore inherently increases the visual complexity of a visualization. When the data are already complex to convey, and many of the more common or accurate visual variables are in use, allocating an additional dimension is non-trivial. As the number of dimensions increases, finding visual variables that are both perceptually accurate (in either estimation of quantity or discrimination of category) as well as perceptually separable from all the other encoding channels, becomes more and more difficult.

A further hurdle is that not all visual variables are well-suited for conveying uncertainty. MacEachren \ea~\cite{maceachren2012visual} evaluate a number of visual variables with respect to their \emph{semiotic fit} for representing uncertainty. They observe that certain visual variables such as blurriness and transparency seem to have a more intuitive connection to uncertainty than other variables such as shape or hue. Unfortunately, Boukhelifa \ea~\cite{boukhelifa2012evaluating} find that many visual variables habitually used for conveying uncertainty, such as blur and value, are also difficult to estimate. This results in a preference/performance gap where designers must choose between encoding uncertainty in a way that is intuitive but error prone, or use higher fidelity channels that may be more difficult to interpret.

Foundational work in psychophysics and graphical perception make use of the concept of a ``just noticeable difference'' (JND)\cite{cleveland1984graphical}. A 50\% JND is a difference between two stimuli that is detectable 50\% of the time. For scales and legends, we would want a significantly higher standard of unambiguity. That is, the different levels of visual variables used in visualizations should be readily and reliably distinguishable. Complicating this goal is that a JND as measured in a controlled environment with cooperative participants can be considerably lower than JNDs measured in the sorts of heterogeneous conditions found in real environments~\cite{szafir2014adapting}.

\subsection{Bivariate Maps}

For spatial visualizations like choropleth maps, heatmaps, and treemaps, visual channels such as position and length are reserved for data variables other than value (such as geographic location or relative size). In these situations, data value is often encoded using color. Colors in such univariate quantitative color maps should be sufficiently far apart as to be perceptually distinguishable~\cite{ware1988color}, and vary in lightness as well as hue to afford an implicit ordering of value~\cite{borland2007rainbow,rogowitz2001blair}. Different choices of color maps can highlight different features of the data, and should be chosen with care based on the data types and important features to be visualized~\cite{rogowitz1996not}.

Given these considerations, and the perceptual integrality of color channels~\cite{garner1970integrality, ware2012information}, the construction of \emph{bivariate} color maps is difficult. Viewers need to be able to independently assess the variables of interest, but the resulting bivariate map should also be expressive and easy to interpret. Bivariate maps with these constraints are often limited to relatively few categories (say, a 3x3 matrix as in \figref{fig:example}) \cite{robertson1986generation,trumbo1981theory}. In general, the quality of bivariate color maps is a multivariate measure involving consideration of not just the component color channels, but also the interpolation scheme and the color of the surround~\cite{bernard2015survey}. However, even simple bivariate maps can be difficult for a general audience to interpret \cite{wainer1980empirical}.

Dunn~\cite{dunn1989dynamic} attempts to circumvent these issues by dynamically allocating bins to bivariate maps. As the number of distinct, separable colors is limited, the dynamic approach allocates bins to regions of the highest utility. For instance, placing the dividing line between bins in areas aligned with a trend line allows the color to encode information about the sign of the residual with respect to each variable. This insight, that the bins in a bivariate map are a limited resource and so should be ``spent'' wisely, drives the creation of VSUMs, which allocate more bins in a bivariate map to regions with less uncertainty.

\section{Value-Suppressing Uncertainty Maps}

Value-Suppressing Uncertainty Maps (VSUMs) are a technique for creating bivariate maps of data \emph{value} and \emph{uncertainty}. These maps circumvent the issues in the creation of traditional bivariate maps by making the following assumptions:

\begin{enumerate}
	\item \textbf{Designers have a limited budget of discriminable marks.} That is, viewers cannot easily or reliably discriminate between an infinite number of categories. Rather than accept this perceptual ambiguity, designers ought to \emph{alias} together marks into a finite, but sufficiently discriminable, set.
	\item \textbf{Uncertain values should have less weight than certain values.} Weight, in this case, might mean literal visual weight, or impact on the decision-making process.
\end{enumerate}

%MC this paragraph is a little iffy, and sort of "the lady doth protest too much" territory. we might be able to get away with "people bin colors all the time. and in fact, you really should be binning your colors anyway."
The first assumption seems to contradict Mackinlay's \emph{expressiveness principle}~\cite{mackinlay1986automating}, where the visual encoding of the data should reflect the composition of the data. For instance, if my data are a collection of measurements of some continuous variable, then it would make sense to encode these data using a continuous \emph{visual} variable, which would then mean an arbitrary number of potential marks. For instance, a bar can have an arbitrary, non-integer height. However, for many visualizations, the perceptual ambiguity caused by having marks that cannot be reliably disambiguated dominates the gains from having a visual encoding that maps readily to the backing data. Padilla \ea~\cite{padilla2017evaluating} present one such example, where binned heatmaps result in higher task accuracy than continuous heatmaps.

In some cases, this second assumption is violated. For instance, an analyst might be interested in ``long tail risks'' or other ``black swan events,'' where the impact of a value, no matter how uncertain, must be considered and planned for~\cite{taleb2011black}. Other analysis tasks (such as filtering out outliers), require increased, rather than decreased, discriminability when uncertainty is high. However, for many information fusion tasks, the assumption is that uncertainty is related to data quality, or the uncertainty of data values~\cite{riveiro2007evaluation}. As with null hypothesis significance testing, the aim would then be to avoid Type I errors, where patterns arising from statistical noise or measurement error are interpreted as significant signals.

If both of these assumptions hold, then it follows that the designer of a bivariate map should allocate more mark types to certain values, and fewer mark types to uncertain values. VSUMs codify this decision by \emph{reducing the number of mark categories for representing value as uncertainty increases.} This means that data encoded using a VSUM will make visible only the largest of differences in uncertain data, but highlight comparatively small differences in value when uncertainty is low. VSUMs therefore act as both a filtering mechanism (in that values with too much uncertainty are all mapped to the same glyph), as well as an implicit test of effect size (in that smaller and smaller changes in value are visible as uncertainty decreases). This strategy of dampening the low quality or high uncertainty values in maps in order to focus on more informative regions has measurable benefits, including the removal of statistically spurious visual patterns, and the highlighting of regions of interest~\cite{correll2017surprise}.

\subsection{Method}

\flowFig

The general VSUM construction algorithm is as follows:

\begin{enumerate}
	\item \textbf{Select a visual variable to encode value}. In this paper, this variable is position along some established color scale.
	\item \textbf{Select a visual variable to encode uncertainty}. We prefer variables that have a somewhat high semiotic connection with uncertainty (such as lightness and size), but still have a reasonable number of perceptually discriminable layers (as opposed to other, more intuitive channels such as blur or sketchiness). In most cases in this paper, we encode uncertainty as interpolation to white or as glyph size.
	\item \textbf{Select a discriminability threshold}. Choose some minimum distance, in a perceptual space, between marks. For many visual variables (such as length or hue), the perceptual discriminability between marks has been assessed in prior literature, and so the only step is to choose a value in terms of the just noticeable distance (JND). For less common visual variables, a perceptual model may be built empirically (as in Demiralp \ea~\cite{demiralp2014learning}).
	\item \textbf{Choose how quickly categories \emph{degrade}}. That is, if there are $n$ bins at some uncertainty level $m$, how many bins should there be when uncertainty increases to level $m+1$? In \figref{fig:example}, this function is $\left \lfloor n / 2 \right \rfloor$, meaning that the number of value bins decreases by half for each level of uncertainty bins. If there is too much degradation, then there are fewer levels of uncertainty. If there is no degradation, the resulting map is just a traditional bivariate map.
	\item \textbf{Choose the largest number of initial categories $n$ such that the discriminability threshold is met.} For each pair of marks in the map, perform a pairwise comparison of discriminability. If no marks are perceptually ambiguous by the standards of your threshold, the map is complete. Otherwise, decrement $n$.
\end{enumerate}

\figref{fig:flow} shows this process in detail, and how it can be used to generate a bivariate map for presenting uncertainty. JavaScript code for generating VSUMs given arbitrary D3~\cite{bostock2011d3} quantitative scale functions is available at \url{URL-REMOVED-FOR-REVIEW}.

Correll \ea~\cite{correll2015layercake,correll2011visualizing} use a precursor of VSUMs in their LayerCake genomics visualization tool, where marks representing genomic data with increasing uncertainty are mapped to a smaller and smaller set of increasingly grey colors, creating the effect of uncertain values retreating into ``confidence fog'' while highly certain values remain prominent. Other bivariate visualizations implicitly alias together uncertain values. For instance, if uncertainty is encoded by transparency, a maximally uncertain glyph may be entirely transparent, and so impossible to distinguish from any other maximally uncertain glyph. Other channels with a semiotic connection to uncertainty, such as saturation, value, blur, or size, also have deleterious effects on the disambiguation of colors and shapes. In both cases, the property of aliasing is \emph{ad hoc}, and places no guarantees on the discriminability of colors. The binning and degradation approach of VSUMs makes the choice to alias values explicit to both the designer and the viewer, and results in a bivariate mapping with known perceptual properties.

\subsection{Design Considerations}

\performanceFig

There are multiple choices that designers must make before creating a VSUM. In particular, they must choose 1) which visual channels map to value and uncertainty, 2) a threshold of perceptual distinguishability (and, implicitly, a corresponding metric), and 3) a degradation function to determine how quickly uncertain values are aliased.

Often, the graphical conventions of a chart reserve variables such as position and area for other uses, and determine the visual channel to be used for value. In heatmaps, for instance, value is regularly encoded using color, while vertical and horizontal position are reserved for location. In this paper, we primarily deal with the case where value is encoded using color. While there are some similar conventions for representing uncertainty, generally designers have more freedom in choosing a visual channel for uncertainty~\cite{maceachren1992visualizing}. We recommend channels with both a strong semiotic connection to uncertainty as well as a relatively large number of perceptually distinguishable levels. Here, we focus on value/saturation (or, equivalently, transparency against a white background) and size. Correll \& Gleicher~\cite{correll2013error} show that even audiences without statistical backgrounds can interpret uncertainty information encoded in these channels.

Another reason for limiting our discussion of bivariate mappings to color and size channels is that the perceptual distinguishability of these channels has been well-studied by prior work. CIELAB and other color spaces are designed to approximate perceptual uniformity\,---\,that is, distance in the color space should universally correspond with perceptual distance. Recent work~\cite{stone2014engineering,szafir2014adapting} has investigated the discriminability of colors in such spaces not only under laboratory conditions, but also among the disparate display conditions and color vision proficiency of crowdworkers. Similarly, formative work by Cleveland \& McGill~\cite{cleveland1984graphical} assesses the discriminability of marks of different height, and this too has been extended both to a larger population of crowdworkers, and to the area channel specifically~\cite{heer2010crowdsourcing,talbot2014four}.

We use values from these works as benchmarks for selecting thresholds of discriminability. The JNDs of these channels provides a useful minimum; in practice, we would wish to create maps with marks that are far more reliably distinguishable than the 50\% or 75\% discriminability rate of a JND. \figref{fig:performance} shows how these benchmarks can be used to select reasonable thresholds. Other factors, such as nameability and aesthetic preference, can be added in as additional weights impacting the desirability of certain color palettes, as in Gramazio \ea~\cite{gramazio2017colorgorical}.

How quickly the number of bins decreases is another parameter of VSUMs. If this degradation is too fast, then we may only have a few discriminable levels of uncertainty. If this degradation is too slow, then we can be overwhelmed with colors. For instance, if we only decrement by one bin per level of uncertainty, there would be $1,485$ different colors in a VSUM with a tolerance threshold of $1$ unit in CIELAB space, far too much complexity for a discrete color mapping. In the examples in this paper, we halve the number of bins at each increasing level of uncertainty. When the number of initial bins at the highly certain region is a power of $2$ (as in Fig. \ref{fig:example}), this imparts a tree structure to the resulting VSUM. As uncertainty increases, bins in the previous level map to one and only one ``parent'' in the next uncertainty level, resulting in fewer crossings of color boundaries among rows.

\subsection{Examples}

We present a set of examples showing how VSUMs can be applied to realworld datasets. The decision of VSUMs to allocate color bins asymmetrically amongst value and uncertainty regions affords greater discriminability in regions of the chart that warrant closer scrutiny, while discouraging exploration of regions with noisy or unreliable signals.

\subsubsection{Air Travel Delay}

\airlineFig

Using a public data set from the U.S. Bureau of Transportation Statistics~\cite{bts}, we created a heatmap showing average flight delay for different times of the day and days of the week for U.S. carriers in January 2017. We use standard error $\left(\sigma / \sqrt{n}\right)$ as our measure of uncertainty. Starting from an minimum discriminability threshold of $18$ units in CIELAB space, we created two alternate heatmaps of this delay information. \figref{fig:airline2d} shows a traditional bivariate map, whereas \figref{fig:airlineVsum} shows a VSUM generated under the same constraints.
